---
title: "siteOptiSample"
output: rmarkdown::html_vignette
author: John L. Godlee (johngodlee@gmail.com)
vignette: >
  %\VignetteIndexEntry{siteOptiSample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,   # wider images
  fig.height = 6,  # taller images
  fig.retina = 2   # higher resolution
)
```

```{r setup}
library(siteOptiSample)

library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(terra)
library(sf)
library(ggplot2)
library(tidyterra)
library(scico)
library(GGally)
library(patchwork)

# Set seed
seed_val <- 9238
```

Purpose:

* To what extent are existing plots representative of the broader landscape forest structure? 
* Identify optimal locations for new plots to fill gaps in multidimensional forest structure space.

```{r}
# Import structural metrics raster
data(san_lorenzo_rast)

# Import polygons of existing plots
data(san_lorenzo_plots)

# Import stem data from a single large plot
data(sample_stems)

# Prepare data objects
r <- unwrap(san_lorenzo_rast)
p <- san_lorenzo_plots
s <- sample_stems

```


```{r}
# Visualize correlation among structural layers
ggcorr(as.data.frame(r), label = TRUE)
```

The structural metrics are highly correlated.

# Part 1: Assess current plot representativeness

Extract metrics from raster for each existing plot. `extractPlotMetrics()` returns a dataframe of values in the same row order as `p`. 

If the structural metrics for each plot are already known, you could skip this step.

```{r}
old_ext <- extractPlotMetrics(r, p, fun = mean)
```

Put pixels and existing plots in the same PCA space. `PCALandscape()` returns two objects with PCA scores, one for pixels (`r_pca`) and one for plots (`p_pca`). `r_pca` is a full PCA object, while `p_pca` is only PCA scores. The PCA is run on variables scaled to unit variance shifted to center on zero. In both `r_pca` and `p_pca`, columns are PCA axes and rows are observations.

```{r}
old_pca <- PCALandscape(r, old_ext, center = TRUE, scale. = TRUE)
```

Calculate distance from each pixel to nearest plot in PCA space. Two methods are available: "euclidean" and "mahalanobis". `pcaDist()` returns vector of distances for each pixel in `r_pca`.


```{r}
old_dist_euclidean <- pcaDist(old_pca$r_pca$x, old_pca$p_pca, 
  n_pca = 3, k = 1, method = "euclidean")

old_dist_mahalanobis <- pcaDist(old_pca$r_pca$x, old_pca$p_pca, 
  n_pca = 3, k = 1, method = "mahalanobis")

old_dist_list <- list(
  "euclidean" = old_dist_euclidean,
  "mahalanobis" = old_dist_mahalanobis)
```

Compare Euclidean and Mahalanobis distances:

```{r}
data.frame(
  euclidean = c(old_dist_euclidean), 
  mahalanobis = c(old_dist_mahalanobis)) %>% 
  ggplot(., aes(x = euclidean, y = mahalanobis)) + 
  geom_abline(linetype = 2, colour = "red") + 
  geom_point(alpha = 0.5) + 
  geom_smooth(colour = "blue") + 
  geom_smooth(method = "lm", colour = "green") + 
  theme_classic() +
  labs(
    x = "Euclidean distance",
    y = "Mahalanobis distance") + 
  ggtitle("Distance in structural space from each pixel to nearest plot")
```

```{r}
data.frame(
  euclidean = c(old_dist_euclidean), 
  mahalanobis = c(old_dist_mahalanobis)) %>% 
  pivot_longer(everything()) %>% 
  ggplot(., aes(x = value, fill = name)) + 
    geom_histogram(position = "identity", alpha = 0.5) + 
    scale_fill_discrete(name = "Distance") + 
    theme_classic() + 
    labs(
      x = "Distance",
      y = "Number of pixels")
```


`pcaDist()` can also be run using a matrix of raw variables. The contribution of individual variables to the distance metric can be weighted using the `w` argument in `pcaDist()`, which is a vector of weights for each variable, corresponding to columns in `x` and `y`:

```{r}
# Prepare matrices with raw variables 
vars_raw <- names(r)[c(1,3,5)]
r_vars_raw <- values(r)[,vars_raw]
old_ext_vars_raw <- old_ext[,vars_raw]

# Run distance function with raw variables
old_dist_raw <- pcaDist(r_vars_raw, old_ext_vars_raw,
  n_pca = 3, k = 1, method = "euclidean")

# Run distance function with weighted raw variables
old_dist_raw_weights <- pcaDist(r_vars_raw, old_ext_vars_raw, w = c(1, 5, 10), 
  n_pca = 3, k = 1, method = "euclidean")

# Compare weighted and unweighted distances
data.frame(
  raw = c(old_dist_raw), 
  raw_weights = c(old_dist_raw_weights)) %>% 
  ggplot(., aes(x = raw, y = raw_weights)) + 
  geom_abline(linetype = 2, colour = "red") + 
  geom_point(alpha = 0.5) + 
  geom_smooth(colour = "blue") + 
  geom_smooth(method = "lm", colour = "green") + 
  theme_classic() +
  labs(
    x = "Distance with unweighted raw data",
    y = "Distance with weighted raw data") + 
  ggtitle("Distance in structural space from each pixel to nearest plot")
```

Create map of the dissimilarity of pixels from existing plots using both distance methods:

```{r}
# Prepare rasters with distances
old_dist_r_list <- lapply(old_dist_list, function(x) { 
  out <- r[[1]]
  values(out)[!is.na(values(out))] <- x
  return(out)
})
names(old_dist_r_list) <- names(old_dist_list)

vis_map_list <- lapply(names(old_dist_r_list), function(x) { 
  ggplot() +
    geom_spatraster(data = old_dist_r_list[[x]]) + 
    scale_fill_scico(name = paste0("Relative distance to nearest plot"), 
      palette = "bamako",
      limits = c(0, 8),
      oob = scales::squish) +
    geom_sf(data = p, 
      colour = "#E74B5E", fill = NA) + 
    guides(
      fill = guide_colourbar(
        title.position = "top",
        barwidth = 20,
        barheight = 1)) + 
    theme_bw() +
    ggtitle(x) +
    labs(
      x = NULL, 
      y = NULL) + 
    theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "bottom",
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 12))
})
names(vis_map_list) <- names(old_dist_r_list)

wrap_plots(vis_map_list) + 
  plot_layout(
    nrow = 1, 
    guides = "collect") & 
  theme(legend.position = "bottom")
```

Create a PCA biplot of landscape representativeness:

```{r}
# Extract PCA values from r_pca
pca_pt_r <- as.data.frame(old_pca$r_pca$x)
pca_pt_r$type <- "Landscape value" 

# Extract PCA values from p_pca
pca_pt_p <- as.data.frame(old_pca$p_pca)
pca_pt_p$type <- "Existing plot" 

# Combine PCA values
pca_pt_all <- rbind(pca_pt_r, pca_pt_p)
pca_pt_all$type <- factor(pca_pt_all$type, 
  levels = c("Landscape value", "Existing plot"))

ggplot() +
  geom_point(data = pca_pt_all, 
    aes(x = PC1, y = PC2, 
      size = type, colour = type, alpha = type),
      shape = "circle") + 
  scale_colour_discrete(name = NULL) + 
  scale_size_manual(name = NULL, 
    values = c("Landscape value" = 1, "Existing plot" = 3)) + 
  scale_alpha_manual(name = NULL, 
    values = c("Landscape value" = 1, "Existing plot" = 0.5)) + 
  labs(x = "PC1", y = "PC2") + 
  theme_bw() +
  ggtitle("Structural space coverage") +
  theme(plot.title = element_text(hjust = 0.5))
```

Classify pixels by how well-represented they are by the plots, then create a map of pixel classification. `classifRepres()` returns a dataframe with the distances of each pixel in `r_pca` to its nearest plot in structural space, and its classification as well-represented or poorly represented by the plots, according to the confidence interval threshold provided in argument `ci`. 

```{r}
# Run pixel classification
old_pca_classif <- classifRepres(
  r_pca = old_pca$r_pca$x, 
  p = old_pca$p_pca, 
  n_pca = 3, 
  ci = 0.95
)

# Prepare raster with pixel classifications
r_classif <- r[[1]]
values(r_classif)[!is.na(values(r_classif))] <- old_pca_classif$group
r_classif_cls <- data.frame(
  id = c(1, 2), 
  classif = c("Poorly represented", "Well-represented by existing plots"))
levels(r_classif) <- r_classif_cls

# Create map of pixels and their classification
ggplot() +
  geom_spatraster(data = r_classif) + 
  scale_fill_discrete(name = NULL, guide = guide_legend(nrow = 3),
    na.value = NA, na.translate = FALSE) +
  theme_bw() +
  ggtitle("Landscape representativeness") +
  labs(
    x = NULL, 
    y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12))
```

Create a PCA biplot of pixel classification:

```{r}
old_pca_classif_df <- as.data.frame(old_pca$r_pca$x)
old_pca_classif_df$group <- old_pca_classif$group

ggplot(old_pca_classif_df, aes(PC1, PC2, colour = group)) +
  geom_point(alpha = 0.8) +
  scale_colour_discrete(name = NULL) +
  theme_bw() + 
  labs(x = "PC1", y = "PC2") + 
  theme(legend.position = "bottom")
```

# Part 2: Optimally locate new plots

The `plotSelect()` function optimally locates plots based on the relative distances of pixels in structural space, optionally taking into account existing plots.

These algorithms are available for locating new plots:

The `meanminSelect()` algorithm aims to minimize the mean distance of all pixels to their nearest plot in PCA space, iteratively placing new plots in pixels that most reduce the average pixel-to-nearest-plot distance. For each candidate pixel, compute how adding a plot to that pixel would decrease the mean pixel-to-nearest-plot distance, then choose the pixel that yields the largest decrease in mean distance. As a result, plots concentrate in common areas structural space.

The `minimaxSelect()` algorithm aims to minimise the maximum distance between proposed plots and existing plots, iteratively placing new plots in pixels with structural attributes most dissimilar to existing plots. For each candidate pixel, compute the distance to the nearest existing plot, then choose the pixel with the highest distance value. As a result, plots occupy structural extremes.

The `kmeansSelect()` algorithm aims to capture the structural diversity of the landscape using K-means clustering, which first partitions the available pixels into a number of clusters equal to the number of proposed plots, then identifies the centroid of each cluster and finds the plot closest to that centroid.

The `hypercubeSelect()` algorithm captures the landscape's multi-dimensional structural gradient by matching available candidate pixels to theoretical targets generated from equal-probability quantiles of each metric. This approach ensures that proposed plots are distributed evenly across the entire range of structural conditions, representatively sampling both common and rare feature combinations.

The code below runs `plotSelect()` with a few different options, then compares how they affect plot placement.

```{r}
# Test with existing plots
test1 <- plotSelect(
  r = r,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE,
  n_pca = 3,
  coord = NULL,
  method = meanminSelect
)

test1$type <- "With existing plots"

# Test without existing plots
test2 <- plotSelect(
  r = r,
  p = NULL,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE, 
  n_pca = 3,
  method = meanminSelect)

test2$type <- "Without existing plots"

# Test without plots PCA 
test3 <- plotSelect(
  r = r,
  p = NULL,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = FALSE, 
  n_pca = NULL,
  method = meanminSelect)
test3$type <- "Using raw raster values"

# Test using all columns
test4 <- plotSelect(
  r = r,
  p = NULL,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE, 
  n_pca = 12,
  method = meanminSelect)
test4$type <- "Using all PCA axes"

# Combine tests
test_1_2_3_4 <- list(
  test1,
  test2, 
  test3,
  test4)

# Create a raster to use as a background
r_bg <- r[[1]]
r_bg[!is.na(values(r_bg))] <- 1
r_bg <- as.factor(r_bg)

# Compare outputs
wrap_plots(lapply(test_1_2_3_4, function(x) {
  ggplot() + 
    geom_spatraster(data = r_bg, show.legend = FALSE) +
    scale_fill_manual(values = "grey", na.value = NA, na.translate = FALSE) + 
    geom_sf(data = x, fill = NA, colour = "red") + 
    theme_void() + 
    ggtitle(unique(x$type))
}))
```

A mask layer provided by argument `r_mask` can be used to restrict the location of new plots. You could use this to filter out inaccessible locations, for example.

```{r}
# Test with mask layer
r_mask <- r[[1]]
r_mask[values(r_mask) < 35] <- NA_real_
r_mask[!is.na(values(r_mask))] <- 1
r_mask <- as.factor(r_mask)

test5 <- plotSelect(
  r = r,
  p = NULL,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = r_mask,
  pca = TRUE, 
  n_pca = 3,
  method = meanminSelect)
test5$type <- "Using a mask layer"

# Should always overlap mask layer   
ggplot() + 
  geom_spatraster(data = r_mask, show.legend = FALSE) + 
  scale_fill_manual(values = "grey", na.value = NA, na.translate = FALSE) + 
  geom_sf(data = test5, colour = "red", fill = NA) +
  theme_classic()
```

Comparison of the `meanminSelect()`, `minimaxSelect()`, and `kmeansSelect()` algorithms:

```{r}
# Test with minimax algorithm
test6 <- plotSelect(
  r = r,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE, 
  n_pca = 3,
  method = minimaxSelect)
test6$type <- "minimax"

# Test with the kmeans algorithm
test7 <- plotSelect(
  r = r,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  method = kmeansSelect)
test7$type <- "kmeans"

# Test with the kmeans algorithm
test8 <- plotSelect(
  r = r,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  method = hypercubeSelect)
test8$type <- "hypercube"

test_1_6_7_8 <- list(
  test1, 
  test6,
  test7,
  test8)
test_1_6_7_8[[1]]$type <- "meanmin"

wrap_plots(lapply(test_1_6_7_8, function(x) {
  ggplot() + 
    geom_spatraster(data = r_bg, show.legend = FALSE) +
    scale_fill_manual(values = "grey", na.value = NA, na.translate = FALSE) + 
    geom_sf(data = x, fill = NA, colour = "red") + 
    theme_void() + 
    ggtitle(unique(x$type))
}))
```

```{r}
# Compare PCA positions using minimax and meanmin algorithms

# Extract metrics from raster for each existing plot
test1_ext <- extractPlotMetrics(r, test1, fun = NULL)
test6_ext <- extractPlotMetrics(r, test6, fun = NULL)
test7_ext <- extractPlotMetrics(r, test7, fun = NULL)
test8_ext <- extractPlotMetrics(r, test8, fun = NULL)
# Using FUN = NULL returns the values of all values within each pixel

# Put pixels and existing plots in the same PCA space
test1_pca <- as.data.frame(PCALandscape(r, test1_ext, center = TRUE, scale. = TRUE)$p_pca)
test1_pca$type <- "meanmin"
test6_pca <- as.data.frame(PCALandscape(r, test6_ext, center = TRUE, scale. = TRUE)$p_pca)
test6_pca$type <- "minimax"
test7_pca <- as.data.frame(PCALandscape(r, test7_ext, center = TRUE, scale. = TRUE)$p_pca)
test7_pca$type <- "kmeans"
test8_pca <- as.data.frame(PCALandscape(r, test8_ext, center = TRUE, scale. = TRUE)$p_pca)
test8_pca$type <- "hypercube"

pca_pt_test_1_6_7_8 <- rbind(pca_pt_r, test1_pca, test6_pca, test7_pca, test8_pca, pca_pt_p)
pca_pt_test_1_6_7_8$type <- factor(pca_pt_test_1_6_7_8$type,
  levels = c("Landscape value", "meanmin", "minimax", "kmeans", "hypercube", "Existing plot"))

ggplot() +
  geom_point(data = pca_pt_test_1_6_7_8, 
    aes(x = PC1, y = PC2, 
      size = type, colour = type, alpha = type),
      shape = "circle") + 
  scale_colour_discrete(name = NULL) + 
  scale_size_manual(name = NULL, 
    values = c("Existing plot" = 3, "meanmin" = 3, "minimax" = 3, 
      "kmeans" = 3, "hypercube" = 3, "Landscape value" = 1)) + 
  scale_alpha_manual(name = NULL, 
    values = c("Existing plot" = 0.5, "meanmin" = 0.5, "minimax" = 0.5, 
      "kmeans" = 0.5, "hypercube" = 0.5, "Landscape value" = 1)) + 
  labs(x = "PC1", y = "PC2") + 
  theme_bw() +
  ggtitle("Structural space coverage") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Calculate distance from each pixel to nearest plot in PCA space
# Using meanmin and minimax algorithms
test1_dist_euclidean <- pcaDist(old_pca$r_pca$x, 
  as.matrix(test1_pca[,grepl("PC", colnames(test1_pca))]), 
  n_pca = 3, k = 1, method = "euclidean")

test6_dist_euclidean <- pcaDist(old_pca$r_pca$x, 
  as.matrix(test6_pca[,grepl("PC", colnames(test6_pca))]), 
  n_pca = 3, k = 1, method = "euclidean")

test7_dist_euclidean <- pcaDist(old_pca$r_pca$x, 
  as.matrix(test7_pca[,grepl("PC", colnames(test7_pca))]), 
  n_pca = 3, k = 1, method = "euclidean")

test8_dist_euclidean <- pcaDist(old_pca$r_pca$x, 
  as.matrix(test8_pca[,grepl("PC", colnames(test8_pca))]), 
  n_pca = 3, k = 1, method = "euclidean")
 
# Create histograms to compare resulting distances
data.frame(
  existing = c(old_dist_euclidean), 
  meanmin = c(test1_dist_euclidean),
  minimax = c(test6_dist_euclidean),
  kmeans = c(test7_dist_euclidean),
  hypercube = c(test8_dist_euclidean)
  ) %>% 
  pivot_longer(everything()) %>% 
  mutate(name = factor(name, 
    levels = c("existing", "meanmin", "minimax", "kmeans", "hypercube"),
    labels = c("Existing plot", "meanmin", "minimax", "kmeans", "hypercube"))) %>% 
  ggplot(., aes(x = value, fill = name)) + 
    geom_histogram(position = "identity", colour = "black") + 
    facet_wrap(~name, scales = "fixed", ncol = 1) + 
    theme_classic() + 
    theme(legend.position = "none") + 
    labs(
      x = "Distance",
      y = "Number of pixels")
```


The `classifVeg()` function can be used to classify pixels into discrete
vegetation types according to their structural signature. You could then either
stratify the location of new plots among these structural types by splitting
the raster and then using `plotSelect()` like before, or simply use the
vegetation classification as a guide to then locate new plots manually.
`classifVeg()` can run two different clustering algorithms:

* K-means: Groups pixels by minimising the total variance within clusters.
  Tends to create spherical, roughly equal-sized clusters in PCA space.
* Gaussian Mixture Models (GMMs): Fits a mixture of multidimensional
  probability distributions to the data. Unlike K-means, GMM allows clusters to
  have different sizes, shapes, and orientations in PCA space. This makes it
  more effective for preserving rare outlier clusters.


```{r}
# Classify according to K-means with three clusters
classif_kmeans3 <- classifVeg(
  r_pca = old_pca$r_pca$x, 
  k = 3, 
  method = "kmeans", 
  n_pca = 3,
  nstart = 25
)

# Create raster from K-means classification
r_kmeans3 <- r[[1]]
values(r_kmeans3)[!is.na(values(r_kmeans3))] <- classif_kmeans3
r_kmeans3 <- as.factor(r_kmeans3)

# Create plot
ggplot() +
  geom_spatraster(data = r_kmeans3) + 
  scale_fill_discrete(name = "Cluster", na.value = NA, na.translate = FALSE) +
  theme_bw() +
  ggtitle("K-means vegetation clusters") +
  labs(
    x = NULL, 
    y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12))

# Classify according to GMM with three clusters
classif_gmm3 <- classifVeg(
  r_pca = old_pca$r_pca$x, 
  k = 3, 
  method = "gmm", 
  n_pca = 3
)

# Create raster from GMM classification
r_gmm3 <- r[[1]]
values(r_gmm3)[!is.na(values(r_gmm3))] <- classif_gmm3
r_gmm3 <- as.factor(r_gmm3)

# Create plot
ggplot() +
  geom_spatraster(data = r_gmm3) + 
  scale_fill_discrete(name = "Cluster", na.value = NA, na.translate = FALSE) +
  theme_bw() +
  ggtitle("Gaussian Mixture Model vegetation clusters") +
  labs(
    x = NULL, 
    y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12))

# Compare cluster attribution between K-means and GMM
table("Kmeans" = values(r_kmeans3), "GMM" = values(r_gmm3))
```

To avoid an extremely patchy mosaic of vegetation classes, it is advisable to
`aggregate()` fine spatial resolution rasters to a lower resolution. You can
also use the `spatial_weight` argument with spatial coordinates (`coords`) to
encourage spatial clustering. 

```{r}
# Aggregate 50x50 m raster to 100x100 m
r_agg100 <- aggregate(r, 2, method = "mean")
r_agg100_pca <- PCALandscape(r, center = TRUE, scale. = TRUE)

# Classify according to K-means with three clusters
classif_kmeans3_agg100 <- classifVeg(
  r_pca = old_pca$r_pca$x, 
  k = 3, 
  method = "kmeans", 
  n_pca = 3,
  nstart = 25
)

# Create raster from K-means classification
r_kmeans3_agg100 <- r[[1]]
values(r_kmeans3_agg100)[!is.na(values(r_kmeans3_agg100))] <- classif_kmeans3_agg100
r_kmeans3_agg100 <- as.factor(r_kmeans3_agg100)

# Create plot
ggplot() +
  geom_spatraster(data = r_kmeans3_agg100) + 
  scale_fill_discrete(name = "Cluster", na.value = NA, na.translate = FALSE) +
  theme_bw() +
  ggtitle("K-means vegetation clusters") +
  labs(
    x = NULL, 
    y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12))
```

```{r}
coord_df <- crds(r, na.rm = TRUE)

classif_gmm_spwt_list <- lapply(seq(0, 1, 0.1), function(x) { 
  classif <- classifVeg(
    r_pca = old_pca$r_pca$x, 
    k = 3, 
    method = "kmeans", 
    n_pca = 3,
    spatial_weight = x,
    coords = coord_df,
  )

  r_classif <- r[[1]]
  values(r_classif)[!is.na(values(r_classif))] <- classif
  r_classif <- as.factor(r_classif)

  return(r_classif)
})

wrap_plots(lapply(classif_gmm_spwt_list, function(x) {
  ggplot() +
    geom_spatraster(data = x) + 
    scale_fill_discrete(name = "Cluster", na.value = NA, na.translate = FALSE) +
    theme_bw() +
    labs(
      x = NULL, 
      y = NULL) + 
    theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "bottom",
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 12))
})) + 
  plot_layout(guides = "collect", axes = "collect") & 
  theme(legend.position = "none")
```


# Part 3: Spatial data as a matrix

`plotSelect()`, `PCALandscape()`, `meanminSelect()`, `minimaxSelect()`, `kmeansSelect()`, and `hypercubeSelect()` also work with spatial data as a dataframe or matrix. 

```{r}
# Extract structural metrics as a non-spatial matrix
r_df <- values(r)
valid_rows <- complete.cases(r_df)
r_df <- r_df[valid_rows,]
r_coord <- crds(r)
r_df_all <- cbind(r_df, r_coord)

# Extract existing plots as row indices in r_df
p_vect <- vect(p)
p_vect <- project(p_vect, crs(r))
r_cov <- cells(r, p_vect)[,"cell"]
p_ind <- match(r_cov, which(valid_rows))
p_ind <- p_ind[!is.na(p_ind)]
```

`PCALandscape()` returns identical values whether using a `SpatRaster` or matrix for argument `r`:

```{r}
all(PCALandscape(r_df, old_ext, center = TRUE, scale. = TRUE)$r_pca$x == old_pca$r_pca$x)
```

`plotSelect()` gives similar results using either a raster or a dataframe of structural metrics.  When `r` is a dataframe, the function returns indices of the selected rows from the dataframe of structural metrics.

```{r}
test_df1 <- plotSelect(
  r = r_df_all,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE,
  n_pca = 3,
  coord = c("x", "y"),
  method = meanminSelect
)
test_df1_coord <- as.data.frame(r_coord[test_df1,])

test_r1 <- plotSelect(
  r = r,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE,
  n_pca = 3,
  method = meanminSelect
)

test_df2 <- plotSelect(
  r = r_df_all,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE,
  n_pca = 3,
  coord = c("x", "y"),
  method = minimaxSelect
)
test_df2_coord <- as.data.frame(r_coord[test_df2,])

test_r2 <- plotSelect(
  r = r,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE,
  n_pca = 3,
  method = minimaxSelect
)

set.seed(seed_val)  # Set identical seed to ensure identical results
test_df3 <- plotSelect(
  r = r_df_all,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE,
  n_pca = 3,
  coord = c("x", "y"),
  method = kmeansSelect
)
test_df3_coord <- as.data.frame(r_coord[test_df3,])

set.seed(seed_val)  # Set identical seed to ensure identical results
test_r3 <- plotSelect(
  r = r,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE,
  n_pca = 3,
  method = kmeansSelect
)

set.seed(seed_val)  # Set identical seed to ensure identical results
test_df4 <- plotSelect(
  r = r_df_all,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE,
  n_pca = 3,
  coord = c("x", "y"),
  method = hypercubeSelect
)
test_df4_coord <- as.data.frame(r_coord[test_df4,])

set.seed(seed_val)  # Set identical seed to ensure identical results
test_r4 <- plotSelect(
  r = r,
  p = p,
  n_plots = 10,
  p_new_dim = c(100, 100),
  r_mask = NULL,
  pca = TRUE,
  n_pca = 3,
  method = hypercubeSelect
)

p_df_r1 <- ggplot() +
  geom_spatraster(data = r[[1]]) + 
  scale_fill_scico(name = "Canopy height (zmax)", palette = "bamako", 
    na.value = NA) +
  geom_sf(data = p, fill = NA, colour = "red") + 
  geom_sf(data = test_r1, fill = NA, colour = "blue", linewidth = 1) + 
  geom_point(data = test_df1_coord, aes(x = x, y = y ), shape = 21, colour = "black", fill = "cyan", size = 3) +
  theme_bw() +
  ggtitle("meanminSelect") + 
  labs(
    x = NULL, 
    y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12))
  
p_df_r2 <- ggplot() +
  geom_spatraster(data = r[[1]]) + 
  scale_fill_scico(name = "Canopy height (zmax)", palette = "bamako", 
    na.value = NA) +
  geom_sf(data = p, fill = NA, colour = "red") + 
  geom_sf(data = test_r2, fill = NA, colour = "blue", linewidth = 1) + 
  geom_point(data = test_df2_coord, aes(x = x, y = y ), shape = 21, colour = "black", fill = "cyan", size = 3) +
  theme_bw() +
  ggtitle("minimaxSelect") + 
  labs(
    x = NULL, 
    y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12))

p_df_r3 <- ggplot() +
  geom_spatraster(data = r[[1]]) + 
  scale_fill_scico(name = "Canopy height (zmax)", palette = "bamako", 
    na.value = NA) +
  geom_sf(data = p, fill = NA, colour = "red") + 
  geom_sf(data = test_r3, fill = NA, colour = "blue", linewidth = 1) + 
  geom_point(data = test_df3_coord, aes(x = x, y = y ), shape = 21, colour = "black", fill = "cyan", size = 3) +
  theme_bw() +
  ggtitle("kmeansSelect") + 
  labs(
    x = NULL, 
    y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12))

p_df_r4 <- ggplot() +
  geom_spatraster(data = r[[1]]) + 
  scale_fill_scico(name = "Canopy height (zmax)", palette = "bamako", 
    na.value = NA) +
  geom_sf(data = p, fill = NA, colour = "red") + 
  geom_sf(data = test_r4, fill = NA, colour = "blue", linewidth = 1) + 
  geom_point(data = test_df4_coord, aes(x = x, y = y ), shape = 21, colour = "black", fill = "cyan", size = 3) +
  theme_bw() +
  ggtitle("hypercubeSelect") + 
  labs(
    x = NULL, 
    y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12))

wrap_plots(p_df_r1, p_df_r2, p_df_r3, p_df_r4) + 
  plot_layout(
    nrow = 2,
    guides = "collect")
```

`plotSelect()` also has argument `het_q`, a numeric value between 0 and 1
defining a threshold to filter out internally heterogeneous candidate plot
locations. This prevents plots being placed on sharp structural transitions.
`het_q` represents the maximum quantile for the total focal standard deviation
of metrics within candidate plot footprint, e.g. 0.8 excludes the top 20% most
heterogeneous areas. 

```{r}
het_q_list <- lapply(seq(0.2, 1, 0.2), function(x) { 
  out <- plotSelect(
    r = r,
    p = p,
    n_plots = 10,
    p_new_dim = c(100, 100),
    r_mask = NULL,
    pca = TRUE,
    n_pca = 3,
    het_q = x,
    method = minimaxSelect
  )
  out$het_q <- x
  out
})

het_q_poly <- bind_rows(het_q_list)

ggplot() +
  geom_spatraster(data = r[[1]]) + 
  scale_fill_scico(name = "Canopy height (zmax)", palette = "bamako", 
    na.value = NA) +
  geom_sf(data = p, fill = NA, colour = "red") + 
  geom_sf(data = het_q_poly, aes(colour = het_q), fill = NA, linewidth = 1) + 
  scale_colour_scico(name = "Heterogeneity threshold", palette = "roma") + 
  theme_bw() +
  labs(
    x = NULL, 
    y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12))
```

# Part 4: Selecting 1 ha TLS plots based on stem size distribution

This additional code is somewhat separate to the main package, but is included
here to document one approach for selecting 100x100 m (1 ha) TLS plots for
GEO-TREES sites. 

In GEO-TREES, TLS plots should sample a range of stem sizes and forest
composition. We first divide the plot into three habitat types. We then pick
one plot which represents each habitat. We assume that in old-growth tropical
forest any 1 ha chunk will include sufficient numbers of small stems, so we
focus on sampling the large rare individuals that are responsible for the most
biomass.

In this example, a dataframe of stem measurements is provided for one large
ForestGEO style plot. Each stem has a diameter measurement and XY stem map
coordinates. The plot is 400x400 m (16 ha) and is divided into 20x20 m
quadrats. We will use the 20x20 m quadrats to align 100x100 m (1 ha) TLS
subplots. 

```{r}
# Load additional packages
library(vegan)
library(cluster)
```

First determine the habitat types. The code below shows two different
clustering methods: hierarchical clustering `hclust()` using the "ward.D2"
algorithm, and partitioning around medoids using `cluster::pam()`. We subset
the stem data to only stems larger than 10 cm diameter. 

```{r}
# Subset to stems >= 10 cm diameter. 
# Calculate the basal area of each stem.
# Extract row and column of each 20x20 m quadrat.
s_ge10 <- s %>% 
  filter(diam >= 10) %>% 
  mutate(
    ba = pi * (diam / 2)^2,
    col = as.numeric(str_sub(quadrat, 1, 2)),
    row = as.numeric(str_sub(quadrat, 3, 4)))

# Create a community matrix. Basal area per species (columns) within each quadrat (rows)
s_agg <- aggregate(ba ~ quadrat + species, data = s_ge10, FUN = sum)
mat <- as.data.frame.matrix(xtabs(ba ~ quadrat + species, data = s_agg))

# Hellinger transformation of community matrix
# Bray-Curtis dissimilarity matrix
mat_hell <- decostand(mat, method = "hellinger")
dist_bc <- vegdist(mat_hell, method = "bray")

# Hierarchical clustering method 
hc <- hclust(dist_bc, method = "ward.D2")
hc_clust <- cutree(hc, k = 3)

# Partitioning around medoids method
bc_pam <- pam(dist_bc, k = 3)
pam_clust <- bc_pam$clustering
```

Compare the two clustering approaches:

```{r}
# Compare the two clustering approaches
stopifnot(all(names(hc_clust) == names(pam_clust)))
table(hc_clust, pam_clust)

# Create dataframe of clusters per quadrat 
clust_df <- data.frame(
  quadrat = names(hc_clust),
  hc_clust = as.character(hc_clust),
  pam_clust = as.character(pam_clust)) %>% 
  mutate(
    col = as.numeric(str_sub(quadrat, 1, 2)),
    row = as.numeric(str_sub(quadrat, 3, 4)))

s_clust <- s_ge10 %>% 
  left_join(., clust_df, by = c("quadrat", "col", "row"))

hc_clust_tile <- ggplot(clust_df, aes(x = col, y = row, fill = hc_clust)) + 
  geom_tile() + 
  theme_bw() + 
  coord_equal() + 
  ggtitle("Hierarchical clustering")

pam_clust_tile <- ggplot(clust_df, aes(x = col, y = row, fill = pam_clust)) + 
  geom_tile() + 
  theme_bw() + 
  coord_equal() + 
  ggtitle("Partitioning around medoids")

wrap_plots(hc_clust_tile, pam_clust_tile)
```

Next, we find 100x100 m chunks within the plot that contain a sufficient number
of quadrats of a particular habitat type (>= 50%). 

```{r}
# For each habitat type find 1 ha chunks in moving window with >25% quadrats of that type
window_size <- 5  # 100 / 20 = 5
purity_threshold <- 0.5 # At least 50% 
habitat_types <- unique(clust_df$pam_clust)

valid_chunks <- data.frame()

min_col <- min(clust_df$col, na.rm = TRUE)
max_col <- max(clust_df$col, na.rm = TRUE)
min_row <- min(clust_df$row, na.rm = TRUE)
max_row <- max(clust_df$row, na.rm = TRUE)

# Safety check: prevent the loop from running backwards if the grid is too narrow
if ((max_col - min_col + 1) >= window_size && (max_row - min_row + 1) >= window_size) {
  
  # For each habitat type
  for (hab in habitat_types) {
    # For each window position
    for (c in min_col:(max_col - window_size + 1)) {
      for (r in min_row:(max_row - window_size + 1)) {
        
        window_quads <- subset(clust_df, col >= c & col < (c + window_size) & 
          row >= r & row < (r + window_size))
        
        # Calculate proportion based on pam_clust
        purity <- sum(window_quads$pam_clust == hab, na.rm = TRUE) / (window_size^2)
        
        if (purity >= purity_threshold) {
          valid_chunks <- rbind(valid_chunks, data.frame(
            pam_clust = hab,
            col_start = c, col_end = c + window_size - 1,
            row_start = r, row_end = r + window_size - 1,
            quadrat_count = sum(window_quads$pam_clust == hab, na.rm = TRUE),
            purity = purity
          ))
        }
      }
    }
  }
  
  # Sort results by habitat type, then highest purity
  if (nrow(valid_chunks) > 0) {
    valid_chunks <- valid_chunks[order(valid_chunks$pam_clust, -valid_chunks$purity), ]
    rownames(valid_chunks) <- NULL
  }
  
} else {
  message("Grid is too small for a ", window_size, "x", window_size, " window.")
}

# Add boundary coordinates to the valid_chunks dataframe
valid_chunks$xmin <- valid_chunks$col_start - 0.5
valid_chunks$xmax <- valid_chunks$col_end + 0.5
valid_chunks$ymin <- valid_chunks$row_start - 0.5
valid_chunks$ymax <- valid_chunks$row_end + 0.5
quadrat_size <- 20 
valid_chunks$xmin_m <- (valid_chunks$col_start - 1) * quadrat_size
valid_chunks$xmax_m <- valid_chunks$col_end * quadrat_size
valid_chunks$ymin_m <- (valid_chunks$row_start - 1) * quadrat_size
valid_chunks$ymax_m <- valid_chunks$row_end * quadrat_size
valid_chunks$chunk_id <- 1:nrow(valid_chunks)
```

```{r}
# Plot the distribution of valid chunks for each habitat type
ggplot(s_clust, aes(x = x, y = y)) + 
  geom_point(aes(colour = pam_clust), alpha = 0.1) + 
  geom_rect(data = valid_chunks, 
    inherit.aes = FALSE,
    aes(
      xmin = xmin_m, xmax = xmax_m, 
      ymin = ymin_m, ymax = ymax_m, 
      colour = as.character(pam_clust)), 
    fill = NA, linetype = "dotted") +
  facet_wrap(~pam_clust) + 
  theme_bw() +
  theme(legend.position = "none") + 
  coord_equal() +
  labs(colour = "Habitat Type", x = "X (m)", y = "Y (m)")
```

For each candidate chunk, calculate the 95th percentile of stem diameter, then
pick the chunk with the highest value for each quadrat. This could be changed
to the total basal area or any other metric.

```{r}
# Define stem diameter quantile threshold
q <- 1

# Calculate 95th percentile of stem diameter for each valid chunk
valid_chunks$diam_q <- sapply(seq_len(nrow(valid_chunks)), function(i) {
  c_start <- valid_chunks$col_start[i]
  c_end   <- valid_chunks$col_end[i]
  r_start <- valid_chunks$row_start[i]
  r_end   <- valid_chunks$row_end[i]
  
  # Filter stems that fall inside 5x5 chunk
  chunk_stems <- subset(s_ge10, col >= c_start & col <= c_end & row >= r_start & row <= r_end)
  
  if (nrow(chunk_stems) > 0) {
    # return(quantile(chunk_stems$diam, probs = q, na.rm = TRUE, names = FALSE))
    return(max(chunk_stems$diam, na.rm = TRUE))
  } else {
    return(0)
  }
})

# Sort chunks by largest tree diameter descending. 
# If there is a tie in tree size, use purity, then a random value
valid_chunks$random_tiebreaker <- runif(nrow(valid_chunks))
valid_chunks_sorted <- valid_chunks[
  order(
    -valid_chunks$diam_q, 
    -valid_chunks$purity,
    valid_chunks$random_tiebreaker), ]

# Greedy selection to ensure selected 100x100 m chunks do not overlap
selected_chunks <- data.frame()
habitats_needed <- unique(valid_chunks_sorted$pam_clust)

# Helper function: Returns TRUE if two chunks overlap
check_overlap <- function(box1, box2) {
  overlap_col <- (box1$col_start <= box2$col_end) && (box1$col_end >= box2$col_start)
  overlap_row <- (box1$row_start <= box2$row_end) && (box1$row_end >= box2$row_start)
  return(overlap_col && overlap_row)
}

# Iterate through the sorted list and select the best non-overlapping chunks
for (i in 1:nrow(valid_chunks_sorted)) {
  candidate <- valid_chunks_sorted[i, ]
  
  # Skip if already selected a plot for this habitat type
  if (candidate$pam_clust %in% selected_chunks$pam_clust) { next }
  
  # Check if the candidate overlaps with any chunks already selected
  overlaps <- FALSE
  if (nrow(selected_chunks) > 0) {
    for (j in 1:nrow(selected_chunks)) {
      if (check_overlap(candidate, selected_chunks[j, ])) {
        overlaps <- TRUE
        break
      }
    }
  }
  
  # Otherwise, select the chunk
  if (!overlaps) {
    selected_chunks <- rbind(selected_chunks, candidate)
  }
  
  # Stop early if all habitat types have a plot
  if (nrow(selected_chunks) == length(habitats_needed)) {
    break
  }
}
```

Finally plot out the chosen plots along with the original stem data:

```{r}
s_biggest <- s_clust %>% 
  group_by(pam_clust) %>% 
  filter(diam == max(diam))

# Plot the chosen quadrats with the original stem data
ggplot(s_clust, aes(x = x, y = y)) + 
  geom_point(aes(colour = as.character(pam_clust), size = diam, alpha = diam)) + 
  geom_rect(data = selected_chunks, 
            inherit.aes = FALSE,
            aes(xmin = xmin_m, xmax = xmax_m, 
                ymin = ymin_m, ymax = ymax_m, 
                colour = as.character(pam_clust)),
            fill = NA, 
            linewidth = 1.2, 
            linetype = "solid") + # Switched to solid to signify final selection
  geom_point(data = s_biggest, 
    aes(x = x, y = y, fill = pam_clust), 
    shape = 21, colour = "black") + 
  scale_fill_discrete(name = "Habitat type") + 
  scale_colour_discrete(name = "Habitat type") + 
  scale_radius(name = "Stem diameter (cm)", range = c(1, 10)) + 
  scale_alpha(name = "Stem diameter (cm)", range = c(0.1, 1)) + 
  theme_bw() +
  coord_equal() +
  labs(x = "X (m)", y = "Y (m)")
```

